---
title: "Leader functions and qualities across 59 cultures: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2
bibliography: Leader2.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=F)
source('analyses2.R')
library(pander)
```


<!-- A thematic note: One of our primary goals is to explore how an evolutionary framework can illuminate when we expect to find cultural, behavioral, physiological, etc. variation and then explain (some of) that variation. Part of going "Beyond WEIRD" is getting to theories that can account for population-level variation and predict invariant universals (or, at least, less variable phenotypes). We want to push things beyond the 'butterfly collecting' of simply documenting population-level variation. -->


# Abstract

Leader-follower dynamics are central to much of social life. Processes of leadership and followership observed in traditional societies, especially among small-scale societies, are often quite distinct from those observed among WEIRD societies and their complex organizations. Many researchers have turned to evolutionary theory to integrate diverse perspectives on human leadership and to better understand the diversity of leadership and followership observed across human cultures and across social contexts. Evolutionary theories of leadership, in turn, draw on ethnographic case studies from traditional societies that are thought to more closely resemble the small, face-to-face societies in which humans evolved. Currently, though, there is limited systematic data on the nature of leadership in such societies.

We build on our previous cross-cultural findings to systematically review the nature of leadership across the full range of human cultural diversity. Over 1200 ethnographic texts from the Human Relations Area Files (HRAF) were coded on leader qualities, their functions, and the costs and benefits for both leaders and followers across diverse social contexts. We discovered evidence for both cultural universals in leadership, as well as important variation, especially when compared to leadership in WEIRD societies. Across all contexts, leaders in the ethnographic record are most commonly described as intelligent and knowledgeable, of high social status, accomplished, wealthy, and prosocial. The primary benefits leaders offer to followers are resolving conflicts, organizing cooperation, providing social functions, leading in inter-group conflicts, providing ritual services, and distributing resources. The primary benefits leaders receive from followers include non-food resources, increased social status, and social services from followers. The most frequent costs of leadership for leaders include loss of social status and resources, and increased risk of harm or conflict. For followers, the most frequent costs include loss of resources, costs in providing social services, and increased risk of harm or conflict.

Leadership in WEIRD settings often differs substantially from non-WEIRD settings. In WEIRD settings influence often extends into multiple domains, organizations are more hierarchical, and interactions between leaders and followers are less frequently face-to-face. A detailed view of leaders, their roles, and the costs and benefits of leadership across traditional societies and across contexts stands to make a significant empirical contribution for the development of more integrative evolutionary theories of leadership and social hierarchy.

# Introduction

>It's good to be king, if just for a while. - Tom Petty

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, pressures from competing leaders or rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and cultural forces and group dynamics likely play an out-sized role in contributing to organizational variation across human populations including social and economic trajectories of populations [@wiessner_risk_1982], and cultural evolutionary processes [@henrich_big_2015]. Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010 and @kaplan_evolutionary_2009], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016]. This approach as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leaderhip scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers encur across distinct leadership systems. 

Evolutionay scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with specific rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

We first briefly review current theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership for both leaders and followers across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations.  


<!-- Reference a historical example of a short term leader implementing change with lasting effects, good and/or bad. -->

## The functions of leadership across cultures and contexts

Leadership as a phenomena is a universal feature of human groups; among all societies there are particular contexts in which one individual maintains greater influence. Also, adopting the broad definition of leadership given above, parents can be defined as leaders of nuclear families. Strong community leadership, however, is common cross-culturally [@garfield_review_2019; @lewis_leaders_1974].  

The emergence of leadership and the functions leaders serve across cultures is closely tied to social and political structure. 

Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, their greater knowledge and social support to express their influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. 

Within communities, leaders may emerge whenever groups of individuals form. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also relatively common. 

Increased sedentarization and defensibility of resources is often associated with both social stratification and an institutionalized managerial role of community leaders. Among horticulturalists and agriculturalists managerial leadership can promote for more efficient use of shared resources such as water reserves and cooperative labor. Pastoralists and coastal populations can face similar pressures surrounding grazing lands and fishing access. In the wake of economic intensification, the function of leadership becomes increasing associated with the distribution of resources.

Defensibility of resources is also often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territory are often forced to manage military forces and develop offensive and defensive strategies in warfare, more so than mobile populations.

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies with specific responsibilities and functions associated with their position. The structure of these organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as a figure head embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications. 

The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leadership effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied of industries across 24 countries. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, leaders moderate their behavior through culturally learned values and leadership-specific values.  

The GLOBE CEO Study defines six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, the GLOBE CEO Study found that charismatic leadership behavior is consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and human-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEOs behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. 


## The qualities of leaders across cultures and contexts

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders exist, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations. 

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such physical embodiment might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement. 








## The benefits of leadership across cultures and contexts

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context are able to promote group policies which either align with their individual interests or which are not especially costly for themselves.

Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. 

Institutionalized leadership positions are also typically associated with increased social status. 

Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need. 



## The costs of leadership across cultures and contexts

The unique costs associated with leadership positions have received less attention, relative to the benefits of leadership. 

Group responsibilities necessarily impose opportunity costs on individual priorities and agendas. 

In managing collective action and enacting group-level policies, some behaviors of leaders will upset some followers and leader-follower interactions can be hostile. 




# Study aims

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and context and the functions they serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across from a representative cross-cultural sample. We also aim to understand how leader qualities relate to their functions and how these relationships vary across social contexts and levels of cultural complexity. We hope these empirical results will facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundation component of human social behavior and social organization. 


# Methods

## Ethnographic sample and coding

We evaluate the cross-cultural frequency in the functions and qualities of leaders and the costs and benefits of leadership for both leaders and followers using an ethnographic database previously built and published (doi.org/10.5281/zenodo.2541999), from which @garfield_evolutionary_2019 "tested" three theoretical models against the ethnographic record. Whereas @garfield_evolutionary_2019, documented supporting evidence in reference to influential theoretical models, our goals here are to let the ethnographic record speak for itself in understanding variation in the dimensions of leadership we are interested in, i.e., our methods and analyses are exploratory and not strictly theoretically driven. 

Here we draw on the same database of 1212 ethnographic paragraphs (termed text records) extracted from the 60-culture Probability Sample Files (PSF) of the electronic Human Relations Area Files (eHRAF) database [@naroll_proposed_1967]. The leadershipdata database was designed to capture a wide sample of the ethnographic record of leadership. For the current study, two authors coded each paragraph for presence of `r length(quality_vars)` distinct qualities of leaders, `r length(function_vars)` distinct functions leaders serve, and `r length(leader_benefit_vars)` benefits and `r length(leader_cost_vars)` costs that leaders or followers may receive. 


The operationalized qualities and functions (See Table X) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Tables X. ZG then followed the same process of coding the entire sample of 1212 text records for the X operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository. 

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs or benefits in the following categories: social status/reputation, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, other resources, and other costs or benefits not captured by these categories. See Table X. 

For additional details on ethnographic sampling and creating of the leadershipdata package see CITE. 

## Statistical analysis

We analyzed data from our researcher-coded operationalized variables at the text-record level and at the culture-level. Our first goal was to assess the proportion of text records and cultures that provided evidence for each of our coded variables to discover the qualities, functions, benefits and costs of leadership that are most frequently noted in the ethnographic record.  

<!-- At the text record level, hierarchical cluster analysis was then used to identify natural groupings of all our coded variables. If variables from a model clustered together, and separately from the other variables, that would lend support to the theoretical cohesiveness of that model. If, on the other hand, variables from different models clustered together, that would suggest theoretical models may need revision or that clusters had identified components of a novel theoretical model.  -->


To reduce the dimensionality of the `r length(quality_vars)` quality variables and the `r length(function_vars)` function variables, all of which take binary values (0, 1), we used logistic PCA from the Binary Dimensionality Reduction (logisticPCA) package [@logisticPCA]. Prior to analysis we removed all rows that had no evidence for any variable. We optimized rank, $k$, and $m$ (the value to approximate the saturated model) using cross-validation, per recommendation of the package authors. Results from logistic PCA stand to (1) reveal which measures (coded variables) distinguish leaders and (2) produce a quantitiave measure of leader qualities and leader functions for additional exploratory analyses. 


# Results

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic), with cultures labeled by mode of subsistence and represented by a symbol proportional to the number of documents that provide ethnographic texts for that culture (for details, see Table S1). The `r nrow(leader_text2)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(leader_text2$d_culture))` cultures were generally short, with a median word count of X, a mean of X, and standard deviation of X. The vast majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with some records not sex-specific.

```{r geographic, fig.width=15, fig.cap="The geographical distributions of cultures in this analysis. Symbol shapes indicate mode of subsistence and symbol size indicates the number of documents available for that culture."}
library(maps)
world_map <- map_data("world")
p_map <- ggplot(world_map, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group), fill="lightgray", colour = "white") +
  geom_point(
    data = leader_cult, 
    aes(
      c_longitude, 
      c_latitude, 
      colour = subsistence, 
      shape = subsistence, 
      size = documents
    )
  ) +
  coord_fixed() +
  labs(x='', y='', shape = 'Subsistence', colour = 'Subsistence', size = 'Documents') +
  guides(shape = guide_legend(override.aes = list(size = 4))) +
  theme_bw(15)
p_map
# ggsave(filename = 'Figures/map.pdf', p_map)
```


<!-- NEED TO DISCUSS 1187 and split records.  -->

## Variable distributions

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Table X), we computed the percentage of text records that provided support, estimating standard errors using a binomial generalized linear mixed effects model with a random intercept for author nested within culture. See Figure \@ref(fig:varsupport).

```{r varsupport, fig.width=15, fig.height=15, fig.cap="Support for each coded variable of leader functions (top) and qualities (bottom). Blue circles: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Red triangles: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
plot.variable.support
```

The most common *functions* leaders serve, which are documented in over 70% of cultures, are providing counsel or direction in follower's affairs, conflict resolution, and organizing various cooperative activities. Leaders also commonly provide social functions, serve as a representative in inter-group interactions, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members, control economic systems, and are responsible for policy development. Notable infrequent functions identified here include controlling group movement and migrations. 

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent. Also, in about 70% of cultures leaders maintained coercive authority over followers. Strategic nepotism, generosity, and having a favorable personality are also common leader qualities, documented in over 50% of cultures. Notable infrequent qualities of leaders in these data include charisma and fairness. 

## Cluster analyses

```{r qualcluster, fig.cap="Cluster analysis of leader quality variables. Distances are $1 - cor$. Ward agglomeration method. p-values computed with 10,000 bootstrap samples using the pvclust package [@pvclust2015]."}
plot(m_pvclust_qual)
pvrect(m_pvclust_qual, alpha = 0.9)
```

```{r funcluster, fig.cap="Cluster analysis of leader function variables. Distances are $1 - cor$. Ward agglomeration method. p-values computed with 10,000 bootstrap samples using the pvclust package [@pvclust2015]."}
plot(m_pvclust_fun)
pvrect(m_pvclust_fun, alpha = 0.9)
```

## Principle components analyses

A logistic PCA on the `r length(quality_vars)` coded leader quality variables, with rank $k=`r dim(logpca_model_qualities$U)[2]`$ optimized by cross-validation, explained `r signif(logpca_model_qualities$prop_deviance_expl, 3)*100`% of the deviance. We draw on the first of the three leader quality components for interpretation and additional analyses and identify this component as a *Prestige vs. Dominance* measure. Low values on Leader qualities PC 1 (Prestige vs. Dominance) are strongly associated with the variables *Experienced/accomplished* and *Knowledgeable/intelligence* and moderately associated with *Oratory skill* and *Innovative*. High values are strongly associated with *Polygnous* and *Coercive authority* and moderately associated with *Many children* and *High quality spouse*. Therefore, the "dominance" side of the *Prestige vs. Dominance* component is also closely liked to reproduction. Measures of physical formidability and aggressiveness are also moderately associated with the "dominance" side of this component. See Figure \@ref(fig:qualitiesPCplot).  

```{r qualitiesPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader qualitiy variables on the first two components."}
logisticPCA_loadings_plot(logpca_model_qualities, data = pca_data_qualities2)
```

Similarly, A logistic PCA on the `r length(quality_vars)` coded leader quality variables, with rank $k=`r dim(logpca_model_functions$U)[2]`$ optimized by cross-validation, `r signif(logpca_model_functions$prop_deviance_expl, 3)*100`% of the deviance. The first Leader functions component distinguishes between interpersonal mediation functions and group organizational and social functions, we define as a *Mediation vs. Organization* measures. Low values on Leader functions PC 1 (Mediation vs. Organization) are strongly associated with the variables *Resolve conflict* and *Provide counsel/direction* and moderately associated with *Political appointments* and *Punishment*, whereas high values are strongly associated with *Social functions* and *Organize cooperation* and moderately associated with *Ritual functions*. See Figure \@ref(fig:functionsPCplot).


```{r functionsPCplot, fig.width=15, fig.height=10, fig.cap="Logistic PCA loadings of coded leader function variables on the first two components."}
logisticPCA_loadings_plot(logpca_model_functions, data = pca_data_functions2)
```

## Benefits and costs of leadership and followership

Leaders are most commonly receiving benefits in the form of greater access to resources and increased social status. Recieving social services and mating benefits are also relatively frequent benefits of leadership cross-culturally. The costs follow the potential benefits and leaders stand to lose their high social status and greater access to resources. Followers on the other hand most commonly benefit from leadership systems through receiving resources and social services. Mating and fitness benefits are relatively rare for followers. Followers most frequently encur costs in the form of resources lost and in providing social services. For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

```{r varsupportcb, echo=FALSE, warning=FALSE, message=F, fig.width=15, fig.height=11, fig.cap="Support for each coded variable for the costs and benefits of leadership and followership. Blue circles: percent of text records with evidence of support (95% CI computed with a binomial mixed effects models with random intercept for author nested within culture). Red triangles: percent of cultures with evidence of support (95% CI computed with a cluster bootstrap)."}
plot.variable.support_costs_benefits
```

## Exploring population-level variation

To explore potential population variation in the *Prestige vs. Dominance* leader qualities component and the *Mediation vs. Organization* leader functions component, we fit a series of hierarchical linear models at the text-level, using nested random effects of document within culture with these componets as outcome measures. 

For predictor variables we incorporate additional population-level measures. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “huntergatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the eHRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed. 

We drew on the Standard Cross-Cultural Sample (SCCS) database [@murdock_standard_1969] for additional population-level measures, which scores 186 cultures on many qualitative or quantitative variables, such as kinship system, dependence on agriculture, and social stratification. Fifty of these cultures are the same as, or very similar to, cultures in the PSF. Specifically, we use the SCCS measures of community size, population density, and cultural complexity. Using these measures as covariates allows us to determine the degree to which, an aggregate measure of cultural complexity, mode of subsistence, or measures of community size and population dispersal most strongly impact the qualities and functions of leaders. We also include our researcher coded measure of group structure (See Table X) as a predictor.  


```{r, fig.height=20, fig.width=9}
# qc_bayes_model_post_plot
```

# Discussion

# References


